High-level cognition is supported by at least second order dynamic correlations in neural activity patterns

### 2019-08-07 paper/boneyard.tex
re-worked abstract and intro, updated bibliography

>>>>

To understand the neuronal computations that support cognition, we
must understand the cooperative dynamics of populations of neurons.
These populations of neurons interact within each brain structure, and
the structures interact to form complex and dynamic networks. These
interactions, at each scale, vary according to the functions our
brains are carrying out and recent work has shown these dynamic
complex patterns support consciousness~\citep{DemeEtal19}.

In the last several decades, advances in functional magnetic resonance
image (fMRI) analyses have evolved to better characterize these
interactions of brain structures.  As~\cite{Turk13} outlines,
analyses using multivariate patterns of activity have given an advance
over univariate activity patterns because they allow relative
contributions of voxels to be combined and better assess distrubted
representations. Resting-state connectivity (RS) fMRI, which is the
temporal correlation of regions during rest, has shed light on the
rich and complex spatiotemporal organization of spontaneous brain
activity.  Functional connectivity (FC) has further characterized the
cognitive state dependency of this network organization.

Recent work has shown that FC fluctuates over time~\citep{ChanGlov10}~\citep[for review]{LuriEtal18}, and the assumed stationarity of these analyses may be too
simple to capture the dynamic nature of brain activity.  Additionally,
recent work has shown that temporal
variability in functional connectivity predicts attention task
performance~\citep{FongEtal19} and that
dynamic correlations between multivariate voxel patterns can add an
additional boost compared to static multivariate voxel patterns alone~\citep{MannEtal18}.

### 2019-08-07 paper/main.tex
re-worked abstract and intro, updated bibliography

\documentclass[english]{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{apacite}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{pxfonts}
\usepackage[utf8]{inputenc}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage[left]{lineno}
\usepackage{soul}
\linenumbers

\title{High-order dynamic neural correlations reflect naturalistic processing
in humans} \author{Lucy
  L. W. Owen$^1$, Thomas H. Chang$^{1,2}$, and\
  Jeremy R. Manning\textsuperscript{$1, \dagger$}\\
  [0.1in]$^1$Department of Psychological and Brain
  Sciences,\\Dartmouth
  College, Hanover, NH\\
  $^3$Amazon.com, Seattle, WA\\
  \textsuperscript{$\dagger$}Address correspondence to
  jeremy.r.manning@dartmouth.edu}

@@

\begin{abstract}
  Our thoughts arise from coordinated activity patterns across our
  brain. We examined high-order dynamic correlations in functional
  neuroimaging data collect as human participants listened to
  different auditory stimuli varying in cognitive richness, along with
  an additional resting state condition. Our approach combines a
  kernel-based method for estimating the dynamic functional
  correlations that are similar (within task) across participants,
  along with a dimensionality reduction approach that enables us to
  efficiently compute high-order correlations in the data. We trained
  classifiers to decode the precise time, relative to the start of the
  stimulus, when a given neural pattern was recorded. We trained
  these classifiers using the neural activity timeseries, first-order
  dynamic correlations, and higher-order correlations (up to
  tenth-order correlations), and asked which types of features led to
  the highest decoding accuracy. We found that higher-order
  correlations consistently yielded higher decoding accuracy for
  the most cogntively rich condition of the experiment, whereas
  first-order correlations or raw data yielded the highest decoding accuracy at
   for the less cogntively rich conditions as well as rest.
\end{abstract}

@@

\section*{Introduction}
To understand the neuronal computations that support cognition, we
must understand the cooperative dynamics of populations of neurons.
These populations of neurons interact within each brain structure, and
the structures interact to form complex and dynamic networks. These
interactions, at each scale, vary according to the functions our
brains are carrying out and recent work has shown these dynamic
complex patterns support consciousness~\citep{DemeEtal19}.

In the last several decades, advances in functional magnetic resonance
image (fMRI) analyses have evolved to better characterize these
interactions of brain structures.  As~\cite{Turk13} outlines,
analyses using multivariate patterns of activity have given an advance
over univariate activity patterns because they allow relative
contributions of voxels to be combined and better assess distrubted
representations. Resting-state connectivity (RS) fMRI, which is the
temporal correlation of regions during rest, has shed light on the
rich and complex spatiotemporal organization of spontaneous brain
activity.  Functional connectivity (FC) has further characterized the
cognitive state dependency of this network organization.

Recent work has shown that FC fluctuates over time~\citep{ChanGlov10}~\citep[for review]{LuriEtal18}, and the assumed stationarity of these analyses may be too
simple to capture the dynamic nature of brain activity.  Additionally,
recent work has shown that temporal
variability in functional connectivity predicts attention task
performance~\citep{FongEtal19} and that
dynamic correlations between multivariate voxel patterns can add an
additional boost compared to static multivariate voxel patterns alone~\citep{MannEtal18}.

Following this line of reasoning, we expect higher-order network
dynamics might provide even richer insights into the neural basis of
cognition. We sought to test this hypothesis by developing an approach
to inferring high-order network dynamics from timeseries data,
validating the approach using synthetic data, and then applying the approach to
a neuroimaging dataset comprising data collected as participants
listened to a story varying in cognitive richness~\citep{SimoEtal16}.

>>>>

\documentclass[english]{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{apacite}
\usepackage{hyperref}
\usepackage[sort]{natbib}
\usepackage{pxfonts}
\usepackage[utf8]{inputenc}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage[left]{lineno}
\usepackage{soul}
\linenumbers

\title{High-level cognition is supported by at least second order
  dynamic correlations in neural activity patterns}

\author{Lucy L. W. Owen$^1$, Thomas H. Chang$^{1,2}$, and\
  Jeremy R. Manning\textsuperscript{$1, \dagger$}\\
  [0.1in]$^1$Department of Psychological and Brain
  Sciences,\\Dartmouth
  College, Hanover, NH\\
  $^3$Amazon.com, Seattle, WA\\
  \textsuperscript{$\dagger$}Address correspondence to
  jeremy.r.manning@dartmouth.edu}

@@

\begin{abstract}
  Our thoughts arise from coordinated patterns of interactions between
  brain structures that change with our ongoing experiences.
  High-order dynamic correlations in brain activity patterns reflect
  different subgraphs of the brain's connectome that display
  homologous lower-level dynamic correlations.  We tested the
  hypothesis that high-level cognition is supported by high-order
  dynamic correlations in brain activity patterns.  We developed an
  approach to estimating high-order dynamic correlations in timeseries
  data, and we applied the approach to neuroimaging data collected as
  human participants either listened to a ten-minute story or a
  temporally scrambled version of the story, or underwent a resting
  state scan.  We trained across-participants pattern classifiers to
  decode (in held-out data) when in the session each activity snapshot
  was collected.  We found that classifiers trained to decode from
  high-order dynamic correlations yielded better performance on data
  collected as participants listened to the (unscrambled) story.  By
  contrast, classifiers trained to decode data from scrambled versions
  of the story or during the resting state scan yielded the best
  performance when they were trained using first-order dynamic
  correlations or raw activity patterns.  We suggest that as our
  thoughts become more complex, they are supported by higher-order
  patterns of dynamic network interactions throughout the brain.
\end{abstract}

@@

\section*{Introduction}
A central goal in cognitive neuroscience is to elucidate the
\textit{neural code}: the mapping between (a) mental states or
cognitive representations and (b) neural activity patterns. One means
of testing models of the neural code is to ask how accurately that
model is able to ``translate'' neural activity patterns into known (or
hypothesized) mental states or cognitive
representations~\citep[e.g.,][]{HaxbEtal01, NormEtal06, TongPrat12,
  MitcEtal08a, KamiTong05, NishEtal11, PereEtal18, HuthEtal12,
  HuthEtal16}.  Training decoding models on different types of neural
features can also help to elucidate which specific aspects of neural
activity patterns are informative about cognition-- and, by extension,
which types of neural activity patterns might comprise the neural
code.  For example, prior work has used region of interest analyses to
estimate the anatomical locations of specific neural
representations~\citep[e.g.,][]{EtzeEtal09}, or to compare the
relative contributions to the neural code of multivariate activity
patterns versus patterns of dynamic correlations between neural
activity patterns~\citep[e.g.,][]{MannEtal18, FongEtal19}.  An
emerging theme in this literature is that cognition is mediated by
complex dynamic interactions between brain
structures~\citep{SporHone06, BassEtal06, Turk13, DemeEtal19}.

Studies of the neural code to date have primarily focused on
univariate or multivariate neural patterns~\cite[for review
see][NormEtal06], or (more recently) on patterns of dynamic
first-order correlations~\citep[i.e., interactions between pairs of
brain structures;][]{MannEtal18, FongEtal19}.  We wondered what the
future of this line of work might hold.  For example, is the neural
code mediated by higher-order interactions between brain structures?
Second-order correlations reflect \textit{homologous} patterns of
correlation.  In other words, if the changing patterns of correlations
between two regions, $A$ and $B$, are similar to those between two
other regions, $C$ and $D$, this would be reflected in the
second-order correlations between ($A$--$B$) and ($C$--$D$).  In this
way, second-order correlations identify similarities and differences
between subgraphs of the brain's connectome.  Analogously, third-order
correlations reflect homologies between second-order correlations--
i.e., homologous patterns of homologous interactions between brain
regions.  More generally, higher-order correlations reflect homologies
between patterns of lower-order correlations.  We can then ask: which
``orders'' of interaction are most reflective of high-level cognitive
processes?

Another central question pertains to the extent to which the neural
code is carried by activity patterns that directly reflect ongoing
cognition~\citep[e.g., following][]{HaxbEtal01, NormEtal06}, versus
the dynamic properties of the network structure itself, independent of
specific activity patterns in any given set of regions~\citep[e.g.,
following][]{BassEtal06}.  For example, graph theoretic measures such
as centrality and degree~\citep{BullSpor09} may be used to estimate
how a given brain structure is ``communicating'' with other
structures, independently of the specific neural representations
carried by those structures.  If one considers a brain region's graph
theoretic position in the network (e.g., its eigenvector centrality)
as a dynamic property, one can compare how the positions of different
regions are correlated, and/or how those patterns of correlations change
over time.  We can also compute higher-order patterns in these
correlations to characterize homologous subgraphs in the connectome
that display similar changes in their constituent brain structures'
interactions with the rest of the brain.

To gain insights into the above aspects of the neural code, we
developed a computational framework for estimating dynamic high-order
correlations in timeseries data. This framework provides an important
advance, in that it enables us to examine patterns in higher-order
correlations that are computationally intractable to estimate via
conventional methods.  Given a multivariate timeseries, our framework
provides timepoint-by-timepoint estimates of the first-order
correlations, second-order correlations, and so on (up to tenth-order
correlations in this manuscript).  Our approach combines a
kernel-based method for computing dynamic correlations in timeseries
data with a dimensionality reduction step that projects the resulting
dynamic correlations into a low-dimensional space.  We explored two
dimensionality reduction approaches: principle components
analysis~\citep[PCA;][]{Pear01}, which preserves an approximately invertable
transformation back to the original data; and a second non-invertible
algorithm that explored patterns in eigenvector
centrality~\citep{Land95}.  This latter approach characterizes
correlations between each feature dimension's relative
\textit{position} in the network in favor of the specific activity
histories of different features.

We validated our approach using synthetic data where the underlying
correlations were known.  We then applied our framework to a
neuroimaging dataset collected as 125 participants listened to either
an audio recording of a ten-minute story or a temporally scrambled
version of the story, or underwent a resting state
scan~\citep{SimoEtal16}.  We used a subset of the data to train
across-participant classifiers to decode listening times using a blend
of neural features (comprising neural activity patterns, as well as
different orders of correlations between those patterns that were
inferred using our computational framework).  We found that both the
PCA-based and eigenvector centrality-based approaches yielded neural
patterns that could be used to decode accurately.  Both approaches
also yielded the best decoding accuracy for data collected during
(intact) story listening when high-order (PCA:
second-order; eigenvector centrality: fourth-order) dynamic
correlation patterns were included as features.  When we trained
classifiers on the scambled stories or resting state data, only
lower-order dynamic patterns were informative to the decoders.  Taken
together, our results indicate that high-level cognition is supported
by high-order dynamic patterns of communication between brain
structures.  \textbf{JRM STOPPED HERE}
